\documentclass{beamer}
\usetheme{Boadilla}
\usepackage{listings}
\title{McLint: Static Code Analyzer for MATLAB}
\author{Ismail Badawi}
\date{April 12, 2012}

\begin{document}

\AtBeginSection[]
{
    \begin{frame}{Table of Contents}
        \tableofcontents[currentsection]
    \end{frame}
}
 
\lstset{
    basicstyle=\ttfamily\tiny,
    escapechar=|,
    language=Java,
    showstringspaces=false,
    keywordstyle=\color{purple} 
}                 
 
\begin{frame}
\maketitle
\end{frame}

\begin{frame}{Outline}
\tableofcontents
\end{frame}

\section{Motivation}
\begin{frame}{Thoughts}
\begin{itemize}
    \item Compiler errors are often very helpful.
        \begin{itemize}
        \item They tell you \emph{what} and \emph{where} the problem is.
        \end{itemize}
    \item Compiler warnings are also quite useful if you enable them.
    \item But compilers don't warn about everything!
    \item And sometimes we don't have a compiler...
\end{itemize}
\end{frame}

\begin{frame}{The idea}
Let's write a tool that takes source code and generates compiler warnings.
\begin{itemize}
    \item This variable isn't being used!
    \item This code isn't reachable!
    \item This shouldn't be inside the loop!
\end{itemize}
\end{frame}

\begin{frame}{Static analysis tools}
Tools like this exist for many languages.
\begin{itemize}
\item C has the original {\tt lint} (and many others)
\item Java has {\tt FindBugs} (and many others)
\item JavaScript has {\tt jslint}
\item Python has {\tt pychecker}, {\tt pylint}
\item Even MATLAB has a builtin function {\tt checkcode} for this
\end{itemize}
\end{frame}

\begin{frame}{Static analysis for MATLAB}
Why is this useful for MATLAB?
\begin{itemize}
\item MATLAB is a confusing language.
\item MATLAB is aimed at people who are not experienced programmers.
    \begin{itemize}
        \item Scientists, engineers, students
    \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{MATLAB difficulties}
Why is this harder in MATLAB?
\begin{itemize}
\item MATLAB is a very dynamic language.
    \begin{itemize}
    \item Very weak typing
    \item Features like {\tt eval}, {\tt assignin}, {\tt clear}...
    \item Even something like name resolution is complicated.
    \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{MATLAB builtins}
Why is the builtin MATLAB functionality not enough?
\begin{itemize}
\item It's closed, so we can't see how it works
\item We also can't extend it to write analyses of our own
\end{itemize}
We have a MATLAB frontend (McLab) and an analysis framework (McSAF), so why
not do this ourselves?
\end{frame}

\section{McLint}
\begin{frame}{Goals}
\begin{itemize}
\item Develop an \emph{extensible} framework for analyzing code and
generating messages or warnings about it.
    \begin{itemize}
    \item It should allow for analysis \emph{plugins} to be loaded at runtime.
    \end{itemize}
\item Provide a few important builtin analyses.
\item The end result is a tool that could be useful to MATLAB programmers.
\end{itemize}
\end{frame}

\begin{frame}{Overview}

{\tt mclint file1.m ...}
\begin{enumerate}
\item Parse and generate ASTs for the given code
\item Precompute a few fundamental analyses
    \begin{itemize}
    \item Kind analysis, reaching definitions, live variables
    \end{itemize}
\item Load classes implementing {\tt LintAnalysis} in search path
\item Run analyses one by one, gathering their messages
\item Generate a report of all messages
\end{enumerate}
\end{frame}

\begin{frame}[fragile]{{\tt LintAnalysis}}
\begin{itemize}
\item Every analysis class implements this interface:
\begin{lstlisting}[basicstyle=\ttfamily, lang=java]
public interface LintAnalysis {
    void analyze(Lint lint);
}
\end{lstlisting}
\item In addition to this, it must also expose a constructor taking an
{\tt AnalysisKit}, which is just a class wrapping the AST together with
some precomputed analyses.
\item {\tt Lint} maintains the collection of messages generated. Analyses
call its {\tt report(Message)} method to report messages.
\end{itemize}
\end{frame}

\begin{frame}[fragile]{A sample analysis}
\begin{lstlisting}
public class Shadowing extends DefinitionVisitor implements LintAnalysis {
|\pause|
    private static final String WARNING = 
        "Definition of %s shadows a builtin function or constant.";
|\pause|
    private Lint lint;
    private BuiltinQuery query = Builtin.getBuiltinQuery();
    private Set<String> reported = new HashSet<String>();
|\pause|
    public Shadowing(AnalysisKit kit) {
        super(kit.getAST());
    }

    public void analyze(Lint lint) {
        this.lint = lint;
        this.tree.analyze(this);
    }
|\pause|
    private Message shadow(ASTNode node, String name) {
        return Message.regarding(node, "SHADOW_BUILTIN", String.format(WARNING, name));
    }
|\pause|
    @Override
    public void caseDefinition(Name node) {
        if (reported.contains(node.getID()))
            return;
        if (query.isBuiltin(node.getID())) {
            lint.report(shadow(node, node.getID()));
            reported.add(node.getID());
        }
    }
}
\end{lstlisting}
\end{frame}

\begin{frame}[fragile]{Sample output}
This input ({\tt test.m})
\begin{lstlisting}[basicstyle=\ttfamily]
x = 1:100;
sum = 0;
for i = 1:100
    sum = sum + x(i);
end
\end{lstlisting}
Produces this output (assuming this is the only analysis)
\begin{lstlisting}
test.m [2, 1]: (SHADOW_BUILTIN) Definition of sum shadows a builtin function or constant.
test.m [3, 5]: (SHADOW_BUILTIN) Definition of i shadows a builtin function or constant.
\end{lstlisting}
\end{frame}

\begin{frame}{Implemented so far}
\begin{itemize}
\item Trivial checks
    \begin{itemize}
    \item Loop variables changed inside loop
    \item Statements that print output (i.e. not ending with {\tt ;})
    \item Names that shadow builtin names
    \item Unreachable code
    \end{itemize}
\item Slightly more complicated
    \begin{itemize}
    \item Loop invariant computations
    \item Unused variables
    \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}[fragile]{Loop invariants}
\begin{lstlisting}[basicstyle=\ttfamily]
sum = 0;
for i = 1:100
    sum = sum + i*f(100);
end
\end{lstlisting}
Want to say "consider computing {\tt f(100)} outside the loop". 
\begin{itemize}
\item Reaching definitions!
\item For every expression in a loop body
    \begin{itemize}
    \item Gather all the names used
    \item Check whether any of their reaching definitions are inside the loop. 
        \begin{itemize}
        \item If not, mark as invariant. 
        \item Also mark as invariant if there are reaching definitions
        inside the loop, but with a right hand side previously marked
        invariant.
        \end{itemize}
    \end{itemize}
    \item Repeat until a fixed point is reached.
\end{itemize}
\end{frame}
\section{Results}
\begin{frame}{Results}
\begin{itemize}
\item Can be hard to quantify...
    \begin{itemize}
    \item We made the code 37\% more readable!
    \end{itemize}
\item For those analyses dealing with performance issues, 
can run {\tt mclint}, address all the warnings, then see if performance
improves.
\item A common problem with tools like this is that they end up generating
many many warnings about true but ultimately trivial things.
    \begin{itemize}
    \item How do you measure how useful a warning is?
    \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{Future work}
\begin{itemize}
\item Framework stuff
\begin{itemize}
    \item Improve API, analyses require some boilerplate
    \item Configuration
    \begin{itemize}
        \item Suppress specific warnings
        \item Pass options to specific analyses
    \end{itemize}
    \item Better plugin system
    \begin{itemize}
        \item All sorts of weird classpath issues now...
        \item More sophisticated search paths, e.g. install analyses
            for a certain user or certain project
        \item How to deal with analyses that have dependencies?
    \end{itemize}
    \item Support running analyses on IRs
\end{itemize}
\item Analysis stuff
\begin{itemize}
    \item Growing array analysis
    \item More analyses
\end{itemize}
\item Other stuff
\begin{itemize}
    \item Domain specific language to specify properties to check
\end{itemize}
\end{itemize}
\end{frame}

\end{document}
